Raft Distributed Consensus Algorithm over WebRTC (Web Real-Time Communication)

### 2015-09-08 research_class/paper.tex
Fix bib month data. Figure placement.
Proofread and citations. Update seq diagram.

%\begin{multicols}{2}
%\end{multicols}

@@

\begin{enumerate}
\item Simple WebRTC messaging: implement a small JavaScript application that uses the PeerJS client library and an unmodified PeerJS signaling server. Demonstrate communication over WebRTC DataChannel using two browser instances (or browser tabs).

\item Raft.js over WebRTC: combine the existing Raft.js implementation with the small JavaScript application above to demonstrate Raft over WebRTC. This step uses a statically configured three node cluster (three browser tabs). The PeerJS signaling server is polled to retrieve the list of other clients and the Raft cluster is started when the number of PeerJS clients reaches three.

\item Update Raft.js protocol: update the Raft.js implementation to match the latest description of the Raft protocol (as described in "Consensus", Ongaro dissertation). This includes the implementation of the simplified membership change algorithm (sequential changes) and the separation of the log commit and state machine apply concepts. Additionally, the new Raft.js implementation always starts a new Raft cluster as a cluster with a single node and additional nodes are dynamically added to the cluster (as suggested in section 4.4 of the paper).

\item Dynamic Raft.js over WebRTC: enhance the signaling server with support for multiple sessions/channels to enable separate Raft clusters with the same signaling server. Start each cluster with a single node and automatically add and remove Raft nodes as they come and go.

\item Visual representation: display Raft node state and RPC statistics in addition to the internal Raft log messages.

\item Sessionless/Datagram RPC messages: change Raft RPC requests and response messages so that RPC responses have enough context to be handled separately from RPC requests. Specifically this enables Raft to support datagram (sessionless) network transports like WebRTC.

\item Atomic state machine operations: implement a state machine update operation that only succeeds if no other operation has been applied to the same key since the last time this client read from that key.

\item Automatic client request forwarding: extend the client request RPC to also be sessionless/datagram based. Implement an asynchronous JavaScript function that determines which node is the leader and forwards client request RPCs to the leader. This function must also correlate responses to the client request and invoke the callback for the original request.

\item Chat application: add the ability to send a message to the chat log using the client request forwarding functionality together with atomic state machine operation. Show the current chat history based on the chat entries in the state machine.

\item Server as Raft.js peer (aspirational): enable server nodes (Raft.js running on Node.js) to fully participate as Raft cluster nodes.
\end{enumerate}

Steps 1-9 were successfully implemented for this paper. Step 10 was an
aspirational goal because this functionality is not yet supported in
PeerJS although it is being tracked by issue report:
https://github.com/peers/peerjs/issues/103

@@

\begin{enumerate}
\item Static web service endpoint: this serves the static HTML, CSS and JavaScript that make up the Raft over WebRTC web application.

\item New channel endpoint: when a browser loads this endpoint, a new PeerJS channel is created and the browser is redirected to a URL for loading the main application (via static file service above).  The redirect URL has a query string appended that contains the channel ID. In addition the redirect URL has a fragment identifier (hash/history) that communicates to the web application that this is the first server in the channel.

\item Peer list endpoint: this endpoint takes a channel ID and returns the list of WebRTC peers (browser user agents) that are currently registered in that channel.

Refer to the "4.3 Client" section for a description of how these
endpoints interact.
\end{enumerate}

\subsection{Client}

The Raft over WebRTC testing application is designed as a single web
page (`rtc.html`). The application has different starting behavior
depending on the value of the URL query string (the URL component
following the "?" and up to the "\#") and the fragment identifier (the
URL component after the "\#").

When a new Raft cluster is created, the browser (user agent) connects
to the server new channel endpoint. This endpoint causes the server to
create a new channel for PeerJS signaling. Once this is complete the
server responds to the browser with a redirect message that contains
the new channel identifier in the query parameter and an indication
that this is the very first Raft cluster node in the fragment
identifier.

The browser follows the redirect and loads the static web application
at the static file endpoint provided by the server. This includes the
main application page (`rtc.html`), the core raft algorithm
(`base.js` and `local.js`), the PeerJS client library (`peer.js`), the
actual Raft over WebRTC implementation (`rtc.js`), and a layout
stylesheet (`web/demo.css`).

@@

Subsequent nodes are added to the same Raft cluster by starting a new
browser context (separate browser, browser window, or browser tab) and
loading the same URL (which no longer has the "firstServer" fragment
identifier). There is a convenience link provided in the application
page that opens a new window with a new cluster node.

When the page first loads, a PeerJS signaling connection is
established with the server. Once this connection is established, the
application will be notified when any PeerJS clients connect to or
disconnect from the server. After the PeerJS signaling connection is
established, the client immediately queries the Peer list endpoint to
discover any clients that were connected prior to when this node
connected to the server.

The web application also begins a periodic async polling function
(`addRemoveServersAsync`) when the page loads. When the node is a Raft
leader, the function compares the current active list of PeerJS clients to
the Raft cluster node list. If there are any differences between the
two lists, then any new clients are added to the Raft cluster and
missing clients are removed from the Raft cluster one at a time using
the `addServer` and `removeServer` RPC calls. Servers additions are
prioritized before removals (as recommended in Ongaro 4.4) % TODO/CITE
in order to maximize availability.

When an instance of the web application starts that is not
a "firstServer", the normal Raft timers are not automatically enabled
until the server receives an RPC. The prevents the server from
possibly becoming a leader before it is incorporated into the cluster
by an appendEntries RPC from the "true" leader.

@@

\subsubsection{Client Request Redirect Implementation}

In the Raft over WebRTC implementation each local browser is a client
of the Raft cluster. However in Raft, the leader node is the only node
that adds entries to the log so clients on follower nodes must have
a way to communicate with the leader. This is implemented as pair of
functions named clientRequest and clientRequestResponse. clientRequest
is called by the client with the arguments of a client request RPC and
a callback function to be called once the request is completed. The
clientRequest/clientRequestResponse functions work as follows:

\begin{enumerate}
\item the clientRequestResponse is registered with the local Raft.js
    implementation as the callback function anytime
    clientRequestResponse RPCs are received.
\item the clientRequest function keeps track of the current leader ID.
    If the current leader ID is not set or this node is the leader
    then the client's request is sent directly to the local Raft node.
    If the current leader ID is set then the request is sent to that
    node as a normal RPC request over the WebRTC channel.
\item when the clientRequestResponse function is called and the status
    is "NOT\_LEADER" then the current leader ID being tracked by the
    clientRequest function is set to the leader ID returned in the
    response. Then the original request is re-issued over the normal
    WebRTC channel to the leader.
\item when the clientRequestResponse function is called with
    a successful status, the original callback argument of the
    clientRequest function is called with the results from the RPC
    response.
\end{enumerate}

@@

The chat application extends the basic Raft.js over WebRTC client. The
chat application provides a large text area containing the chat
history and a text entry field below it. When the user clicks on the
\emph{Send} button (or pressed enter), the current node ID is
prepended to the beginning of the text line and it is added to
a pending send queue and the text entry field is blanked. A function
to flush sends runs periodically (currently every 100ms) and does the
following:

\begin{enumerate}
\item If there is line of text that has been sent but not yet
    added (applied) to the state machine or there are no lines of text
    on the pending send queue, then the function does nothing.
\item A text line is removed from the head of the pending send queue
    and stored in \emph{curSend} to indicate that a send is in
    progress. The a client request is made using an atomic
    \emph{seqAppend} command to add the new text to the state machine.
    The \emph{seqAppend} operation attempts to append the text line to
    a list at the \emph{history} key in the state machine.
\item When the atomic \emph{seqAppend} operation completes, it may
    have failed (if another text line was appended first), in which
    case the flush function puts the failed text line back on the
    pending send queue. Whether the send was successful or failed, the
    \emph{curSend} variable is set to nil/null to indicate that no
    send is in progress (the next time the flush function runs it will
    attempt to send again).
\end{enumerate}

The chat application also extends the default set of state machine
operations to add a seqAppend operation. This is an atomic operation
that is implemented as follows:

\begin{enumerate}
\item the seqAppend operation operates on a specially structured
    key/value pair in the state machine map. The value is
    a associative map that contains a counter and list structure that
    contains the actual data. seqAppend operations are sent with both
    a count and then new value to append.
\item When a seqAppend operation is applied to the state machine and
    the count value in the command matches the counter in the state
    machines value, then the new value in the request is added to the
    list and the counter is incremented. If the count value does not
    match then counter then the the operation is not applied. In
    either case, a list is returned that contains two values: whether
    the seqAppend operation succeeded and what the current value of
    the counter.
\item the first time the seqAppend operation is used on a non-existent
    key, the structure described above is created with the counter set
    to 0 and the list structure is created with the single value
    passed in the seqAppend operation.
\item clients that call the seqAppend operation should check the
    status value in the result list. If the status is false, then the
    client should retry the operation using the counter value as the
    count in the operation.
\end{enumerate}

@@

The initial way that the application was tested was by loading the
initial landing page to create a Raft over WebRTC cluster with
a single leader. Here is a screenshot of the browser after the first
node is created:

@@

Then four more Raft nodes were then added to the cluster by clicking
on the embedded "create a node" link in the application. Here is
a screenshot of the browser once four new nodes are added to the
cluster (5 total):

@@

In addition, the full chat application was also tested manually by
loading the initial chat application landing page to create a Raft
over WebRTC cluster with a single leader. Then one or more lines of
text were entered into the text line each followed by clicking the
Send button.  Here is a screenshot of the browser after the first
node/tab of the chat application is created and a line of text was
sent:

@@

Then four more Raft nodes were then added to the cluster by clicking
on the embedded "create more clients" link in the application.
Finally, on a non-leader node more text lines were entered and sent by
clicking the Send button. Here is a screenshot of the browser once
four new nodes/tabs are added to the cluster (5 total) and another
text line was sent:

@@

After every change was made to the cluster all the nodes were checked
to ensure that the following steady state properties were observed:
\begin{enumerate}
\item only a single leader
\item all nodes have the same term and transaction log count
\item all non-leader nodes are receiving `appendEntries` RPCs from the leader
\item all nodes have the same perception of the cluster size
\item the transaction log entries are identical on all nodes
\item the state machine is identical on all nodes
\end{enumerate}

The current leader was then terminated. The other nodes were checked
for the steady state properties. This was repeated until only two
nodes remained in the cluster.

After the five node cluster was brought up, state machine commands
were manually sent to leader to add to the transaction log. This was
done using the JavaScript console as there is not yet a way to do this
via the interface. Again the nodes were checked for the steady state
properties.

Then the application was tested was by loading the initial cluster
creation landing page in a browser tab. Then nodes were randomly added
to the cluster (by clicking on the link embedded in the page) or
removed from the cluster (by closing a tab). After each change the
steady state properties were checked.

@@

The results of the manual testing above showed that Raft over WebRTC
was able to perform leader elections correctly, the log and state
machine was replicated correctly to all nodes, and the safety
properties were maintained. During testing the membership of the
cluster was changed numerous times (both node additions and removals)
and the Raft properties were maintained throughout. In addition,
client interaction was tested although the application does not
provide a friendly interface for doing so.

One limitation of the system is that the if the cluster membership
drops from two nodes to one node then the cluster will become
unavailable without manual intervention. This is because when there
are two nodes in the cluster, both nodes are required for for
a majority vote and when one member is forcible shutdown then the
remaining cluster node will be unable to commit transaction entries
and the cluster will become unavailable. There are several options to
get around this problem:

\begin{enumerate}
\item The node can ask the leader to be removed and then wait for the
  transaction to become effective before it disconnects.  This may not
  always be an option especially in a browser context where the user
  may immediately kill the window/tab without giving the application
  a change to remove itself from the cluster.
\item The signaling server can serve as a tie-breaker. This would allow
  the remaining leader successfully commit the removal of the other
  node. In the case of a partition, the signaling server must only
  vote for one of the nodes.
\item The node with the lower ID could have an implicit tie-breaker vote.
  This would allow one node of the partition to continue in the case
  of a network partition. If a node dies or is killed, this would only
  allow the cluster to continue to operate if the node that dies
  happens to have the higher ID.
\item The system could allow manual user intervention. However, this is
  only likely to work correctly in the case where the user of the last
  remaining node can be certain that the other node will never return
  otherwise they is the risk of data-corruption.
\end{enumerate}

Running a Raft cluster in a browser context leads to some level of
timer instability. The JavaScript execution context is a fully event
driven environment with a single thread of execution for a given
JavaScript context. JavaScript code is run when browser (or Node.js)
events trigger JavaScript code that is registered to that event. In
order to implement logic that runs periodically, JavaScript code
is register as a callback timer using the setTimer, setInterval or
requestAnimationFrame APIs. However, none of these mechanisms is
a precise timer. When a timer fires, the callback code they refer to
is added to the JavaScript event queue and events on the queue are
processed in the order they arrive. This means callbacks may be
delayed by other code that is already running or queued first.

However, there is another browser mechanism that is less well known
but that causes even larger delays in timer code execution. Modern
browsers often have dozens or even hundreds of windows and/or tabs
open simultaneously. In order to reduce CPU load, increase user
responsiveness and improve battery life, most modern browsers will
throttle the JavaScript callback timers for all windows and tabs that
are hidden or have not been directly interacted with for some period
of time. This throttling can cause timers to slow down by an order of
magnitude or more.

In the context of Raft over WebRTC, the slowdown of background page
timers means that whenever the leader node is residing on a page that
is in the background, the propagation of transaction log entries slows
down substantially. This also means that other nodes which are in the
foreground are more likely to have an election timeout timer fire and
they are more likely to become a leader because they are more likely
to be able to send and receive votes quickly than nodes that are in
the background. In this sense the problem is self-correcting because
over the long term because higher performing nodes more are likely to
start and win elections. This also indicates dynamic adjustment of
timeout values may be a fruitful area of further study.

In the context of the full chat application the slowdown of background
pages is even more pronounced because each addition to the chat
history can take several messages to fully propagate the new date.

@@

To simplify the execution of the tests and to make the testing
infrastructure more portable, a docker container environment was used
to encapsulate and execute the SlimerJS environment. The container is
built using \emph{test/Dockerfile} which is in turn derived from
the "fentas/slimerjs" docker image
(https://github.com/fentas/docker/blob/master/slimerjs/Dockerfile).
This custom Dockerfile was created because the upstream version of the
Dockerfile used an older version of XULRunner that did not support
more than 7 WebRTC clients communicating together simultaneously.
% (TODO/CITE)

@@

Three different tests were created to test the Raft over WebRTC
functionality and the chat application. Each of the tests takes
a parameter indicating the number of nodes/pages to launch and test
again. The tests scripts files are:

@@

\begin{enumerate}
\item Explore the use of alternate WebRTC transport modes for relaxing
  ordering constraints and delivery guarantees. Since the Raft
  protocol is tolerant of out-of-order and dropped messages, it is
  possible that Raft over WebRTC could be more efficient with these
  one or both of these constraints relaxed.
\item Test the survivability of the cluster when the signaling server is
  becomes completely unresponsive or just very slow.
\item Explore models for dynamically adjusting Raft timeout values to
  account for changing network and browser conditions.
\item Test the ability of the system to catch up stale nodes by
  disconnecting one or more nodes, run the cluster for some period
  time with transactions created periodically, and then reconnect the stale
  clients.
\item Implement a user interface for sending commands to the transaction
  log and for viewing the current state of the Raft state machine.
\item Explore models for dealing with the case where the cluster goes from
  two nodes down to one node. Can usefulness and availability be
  improved without sacrificing Raft safety constraints.
\item Test the system in more various network environments. Determine the
  reliability of establishing direct browser-to-browser communication.
  Determine if corporate firewalls make this as difficult as is
  implied in some of the documentation and standards texts.
\item Implement and test forwarding client requests rather than default
  redirecting the client to the leader. In other words, allow client
  requests to originate from any node of the cluster and be redirected
  automatically to the leader (rather than the current manual
  process).
\item Perform quantitative and real-world testing to better characterize
  the performance, availability and scalability characteristics of the
  system.
\item Explore models for timing out and automatically removing nodes
  when they become unresponsive without needing to be notified by the
  signaling server that the nodes have disconnected.
\item Test larger scale deployments across multiple systems. Determine the
  factors that affect scalability and characterize scaling limits.
\item Perform wider cross-browser testing including Internet Explorer and
  Opera.
\item Implement Raft log compaction.
\item Explore models that allow for a combination of strict shared state
  and high performance bulk data such as media or object blobs. For
  example, the data with full consensus might contain hashes that to
  the refer to bulk data.
\item Implement a Node.js based Raft over WebRTC application. This would
  allow non-browser contexts to participate in the Raft cluster
  together with the browser clients. PeerJS issue
  [103](https://github.com/peers/peerjs/issues/103) must be resolved
  for this to be feasible.
\item Explore a design where some nodes of the cluster are non-voting and
  are not counted as part of transaction commit quorum. However, these
  nodes would still be included in the log replication algorithm. One
  model this would enable would be a configuration where a small group
  of voting nodes serve as the hub for a larger group of non-voting
  nodes. This may enable higher scaling of the system. It also enables
  interesting applications such as a logging or backup nodes.
\end{enumerate}

>>>>

@@

\begin{enumerate}
\item Simple WebRTC messaging: implement a small JavaScript
    application that uses the PeerJS client library and an unmodified
    PeerJS signaling server. Demonstrate communication over WebRTC
    DataChannel using two browser instances (or browser tabs).
\item Raft.js over WebRTC: combine the existing Raft.js implementation
    with the small JavaScript application above to demonstrate Raft
    over WebRTC. This step uses a statically configured three node
    cluster (three browser tabs). The PeerJS signaling server is
    polled to retrieve the list of other clients and the Raft cluster
    is started when the number of PeerJS clients reaches three.
\item Update Raft.js protocol: update the Raft.js implementation to
    match the latest description of the Raft protocol
    \cite{raft_thesis:ongaro14}. This includes the implementation of
    the simplified membership change algorithm (sequential changes)
    and the separation of the log commit and state machine apply
    concepts.  Additionally, the new Raft.js implementation always
    starts a new Raft cluster as a cluster with a single node and
    additional nodes are dynamically added to the cluster
    \cite[Section~4.4]{raft_thesis:ongaro14}.
\item Dynamic Raft.js over WebRTC: enhance the signaling server with
    support for multiple sessions/channels to enable separate Raft
    clusters with the same signaling server. Start each cluster with
    a single node and automatically add and remove Raft nodes as they
    come and go.
\item Visual representation: display Raft node state and RPC
    statistics in addition to the internal Raft logging messages.
\item Sessionless/Datagram RPC messages: change Raft RPC requests and
    response messages so that RPC responses have enough context to be
    handled separately from RPC requests. Specifically this enables
    Raft to support datagram (sessionless) network transports like
    WebRTC.
\item Atomic state machine operation: implement a state machine
    update operation that only succeeds if no other operation has been
    applied to the same key since the last time this client read from
    that key.
\item Automatic client request forwarding: extend the client request
    RPC to also be sessionless/datagram based. Implement an
    asynchronous JavaScript function that determines which node is the
    leader and forwards client request RPCs to the leader. This
    function must also correlate RPC responses to the client request
    and invoke the callback for the original request.
\item Chat application: add the ability to send a chat message to the
    log using the client request forwarding functionality together
    with atomic state machine operation. Show the current chat history
    based on the chat entries in the state machine.
\item Server as Raft.js peer (aspirational): enable server nodes
    (Raft.js running on Node.js) to fully participate as Raft cluster
    nodes.
\end{enumerate}

Steps 1-9 were successfully implemented for this paper. Step 10 was an
aspirational goal because this functionality is not yet supported in
PeerJS \cite{peerjs:103}.

@@

\begin{enumerate}
\item Static web service endpoint: this serves the static HTML, CSS
    and JavaScript that make up the Raft over WebRTC web application.
\item New channel endpoint: when a browser loads this endpoint, a new
    PeerJS channel is created and the browser is redirected to a URL
    for loading the main application (via static file service above).
    The redirect URL has a query string appended that contains the
    channel ID of the new channel that was just created. In addition
    the redirect URL has a fragment identifier (hash/history) that
    communicates to the web application that this is the first server
    in the channel.
\item Peer list endpoint: this endpoint takes a channel ID and returns
    the list of WebRTC peers (browser user agents) that are currently
    connected to that channel.

Refer to Section~\ref{section:design-client} and
Figure~\ref{fig:raft_rtc_sequence} for a description of how these
endpoints interact.

\subsection{Client}
    \label{section:design-client}

The Raft over WebRTC testing application is designed as a single web
page (\texttt{rtc.html} or \texttt{chat.html}). The application has
different starting behavior depending on the value of the URL query
string (the URL component following the "?" and up to the "\#") and
the fragment identifier (the URL component after the "\#").

When a new Raft cluster is created, the user agent (browser) connects
to the server's new channel endpoint
(Figure~\ref{fig:raft_rtc_sequence}~[1]). This endpoint causes the
server to create a new channel for PeerJS signaling
(Figure~\ref{fig:raft_rtc_sequence}~[2]). Once this is complete the
server responds to the browser with a redirect message
that contains the new channel identifier in the query parameter and an
indication that this is the very first Raft cluster node in the
fragment identifier (Figure~\ref{fig:raft_rtc_sequence}~[3]).

The browser follows the redirect and loads the static web application
at the static file endpoint provided by the server
(Figure~\ref{fig:raft_rtc_sequence}~[4]). This includes the
main application page (\texttt{rtc.html} or \texttt{chat.html}), the
core raft algorithm (\texttt{base.js} and \texttt{local.js}), the
PeerJS client library (\texttt{peer.js}), the actual Raft over WebRTC
implementation (\texttt{rtc.js}), a layout stylesheet
(\texttt{web/demo.css}), and higher-level application code if that is
being loaded (\texttt{chat.js})
(Figure~\ref{fig:raft_rtc_sequence}~[5]).

@@

Subsequent nodes are added to the same Raft cluster by starting a new
browser context (separate browser, browser window, or browser tab) and
loading the same URL without the "firstServer" fragment identifier
(Figure~\ref{fig:raft_rtc_sequence}~[9]). There is a convenience link
provided in the application page that opens a new browser window with
a new cluster node (see \ref{fig:browser_1}).

When each page first loads, a PeerJS signaling connection is
established with the server
(Figure~\ref{fig:raft_rtc_sequence}~[6]~and~[11]). Once this
connection is established, the application will be notified when any
PeerJS clients connect to or disconnect from the server.  After the
PeerJS signaling connection is established, the client immediately
queries the Peer list endpoint
(Figure~\ref{fig:raft_rtc_sequence}~[7]~and~[12]~and~[14]]) to discover
any clients that were connected prior to when this node connected to
the server (Figure~\ref{fig:raft_rtc_sequence}~[8]~and~[13]~and~[15]]).

The web application also begins a periodic async polling function
(\texttt{addRemoveServersAsync}) when the page loads. When the node is a Raft
leader, the function compares the current active list of PeerJS clients to
the Raft cluster membership list. If there are any differences between
the two lists, then any new clients are added to the Raft cluster and
missing clients are removed from the Raft cluster one at a time using
the \texttt{addServer} and \texttt{removeServer} RPC calls. Servers
additions are prioritized before removals 
in order to maximize availability
\cite[Section~4.4]{raft_thesis:ongaro14}.

When an instance of the web application starts that is not
a "firstServer", the normal Raft timers are not enabled until the
server receives its first RPC. The prevents the server from
trying to becoming a leader before it is incorporated into the cluster
by an appendEntries RPC from the true cluster leader.

@@

\subsubsection{Client Request Redirects}

% TODO: rename clientRequest* function to not clash with RPC names
% (maybe raftRequest*?)

In the Raft over WebRqTC implementation each local browser is a client
of the Raft cluster. However in Raft, the leader node is the only node
that adds entries to the log so clients on follower nodes must have
a way to communicate with the leader. This is implemented as a pair of
functions named \texttt{clientRequest} and
\texttt{clientRequestResponse}. The \texttt{clientRequest} function is
called by the client with the arguments of a client request RPC and
a callback function to be called once the request is completed. The
\texttt{clientRequest}/\texttt{clientRequestResponse} functions work
as follows:

\begin{enumerate}
\item the \texttt{clientRequestResponse} is registered with the local
    Raft.js implementation as the callback function anytime
    a \texttt{clientRequestResponse} RPC is received.
\item the \texttt{clientRequest} function keeps track of the current
    leader ID.  If the current leader ID is not set or this node is
    the leader then the client request RPC is sent directly to the
    local Raft node bypassing the network transport layer.  If the
    current leader ID is set then the request is sent to that node as
    a normal RPC request over the WebRTC channel.
\item when the \texttt{clientRequestResponse} function is called and
    the status is "NOT\_LEADER" then the current leader ID being
    tracked by the \texttt{clientRequest} function is set to the
    leader ID specified in the response. Then the original request is
    re-issued over the normal WebRTC channel to the leader.
\item when the \texttt{clientRequestResponse} function is called with
    a successful status, the original callback argument of the
    \texttt{clientRequest} function is called with the results from
    the RPC response.
\end{enumerate}

@@

The chat application extends the basic Raft.js over WebRTC client. The
chat application provides a large text area containing the chat
history and a text entry field below it
(see Figure~\ref{fig:browser_chat_1}). When the user clicks on the
\emph{Send} button (or presses enter), the current node ID is
prepended to the beginning of the text line and it is added to
a pending send queue and the text entry field is blanked. A function
to flush sends runs periodically (currently every 100ms) and does the
following:
% TODO: 100ms above should be less

\begin{enumerate}
\item If there is line of text that has been sent but not yet
    added (applied) to the state machine or there are no lines of text
    on the pending send queue, then the function does nothing.
\item A text line is removed from the head of the pending send queue
    and stored in \emph{curSend} to indicate that a send is in
    progress. Then a client request is made using an atomic
    \emph{seqAppend} command to add the new text to the state machine
    (\emph{seqAppend} is describe below).  The \emph{seqAppend}
    operation attempts to append the text line to a list at the
    \emph{history} key in the state machine.
\item When the atomic \emph{seqAppend} operation completes, it may
    have failed (if another text line was appended first), in which
    case the flush function puts the failed text line back on the
    pending send queue. Whether the send was successful or failed, the
    \emph{curSend} variable is set to nil/null to indicate that no
    send is in progress. The next time the flush function runs it will
    attempt to send again. % TODO: reword
\end{enumerate}

% TODO: normalize use of \texttt and \emph

The chat application also extends the default set of state machine
operations to add a \emph{seqAppend} operation. This is an atomic
operation that is implemented as follows:

\begin{enumerate}
\item the \emph{seqAppend} operation operates on a specially structured
    key/value pair in the state machine map. The value of the pair is
    another associative map that contains a counter and list structure
    where the actual chat lines are accumulated. \emph{seqAppend}
    operations are sent with both a count and then new value to
    append.
\item when a \emph{seqAppend} operation is applied to the state
    machine and the count value in the command matches the counter in
    the state machine's value, then the new line in the request is
    added to the value list and the counter is incremented. If the count
    value does not match the counter then the the operation is not
    applied. In either case, a list is returned in the
    \texttt{clientRequestResponse} that contains two values: a boolean
    indicating whether the \emph{seqAppend} operation succeeded and
    a number indicating the current value of the counter.
\item the first time the \emph{seqAppend} operation is used on a non-existent
    key, the structure described above is created with the counter set
    to 0 and the list structure is created with the single value that
    was passed in the \emph{seqAppend} operation.
\item clients that call the \emph{seqAppend} operation should check the
    status value in the result list. If the status is false, then the
    client should retry the operation. Regardless of success of
    failure, then client should store the current value of the counter
    for the next \emph{seqAppend} call or retry.
\end{enumerate}

@@

The initial way that the application was tested was by loading the
basic application to create a Raft over WebRTC cluster with a single
node acting as leader. Figure~\ref{fig:browser_1} shows a screenshot
of the browser after the first node is created:

@@

Then four more Raft nodes were then added to the cluster by clicking
on the embedded "create more nodes" link in the application.
Figure~\ref{fig:browser_5} shows a screenshot of the browser once four
new nodes are added to the cluster (5 total):

@@

In addition, the full chat application was also tested manually by
loading the initial chat application landing page to create a Raft
over WebRTC cluster with a single leader acting as leader. Then lines
of text were entered into the text area each followed by clicking the
Send button. Figure~\ref{fig:browser_chat_1} shows a screenshot of the
browser after the first node/tab of the chat application is created
and a line of text was sent:

@@

Then four more Raft nodes were then added to the cluster by clicking
on the embedded "create more clients" link in the application.
Finally, on a non-leader node more text lines were entered and sent by
clicking the Send button. Figure~\ref{fig:browser_chat_5} shows
a screenshot of the browser once four new nodes/tabs are added to the
cluster (5 total) and another text line was sent from the fifth node:

@@

After every change was made to the cluster all the nodes were checked
to ensure that the following steady state properties were observed:
\begin{enumerate}
\item only a single leader
\item all nodes have the same term and transaction log count
\item all non-leader nodes are receiving `appendEntries` RPCs from the leader
\item all nodes have the same cluster membership state
\item the transaction log entries are identical on all nodes
\item the state machine is identical on all nodes
\end{enumerate}

The current leader was then terminated. The other nodes were checked
again the steady state properties. This was repeated until only two
nodes remained in the cluster.

Next the five node cluster was brought up again and state machine
commands were sent to leader to add to the transaction log. Again the
nodes were checked for the above steady state properties.

Then the application was tested was by loading the initial cluster
creation landing page in a browser tab. Nodes were randomly added to
the cluster (by clicking on the link embedded in the page) or removed
from the cluster (by closing a tab). After each change the steady
state properties were checked.

@@

The results of the manual testing above showed that Raft over WebRTC
was able to perform leader elections correctly, the log and state
machine was replicated correctly to all nodes, and the safety
properties were maintained. During testing the membership of the
cluster was changed numerous times (both node additions and removals)
and the Raft properties were maintained throughout. In addition,
client interaction was tested to make sure commands were still
properly added to the state machine and transaction log.

One limitation of the Raft protocol is that the if the cluster
membership drops from two nodes to one node then the cluster will
become unavailable without manual intervention. This is because when
there are two nodes in the cluster, both nodes are required for for
a majority vote and when one member is forcible shutdown then the
remaining cluster node will be unable to win elections or even commit
transaction entries and the cluster will become unavailable. There are
several options to get around this problem:

% NOTE: optional section

\begin{enumerate}
\item The node can ask the leader to be removed and then wait for the
  transaction to become effective before it disconnects.  This may not
  always be an option especially in a browser context where the user
  may immediately kill the window/tab without giving the application
  a chance to remove itself from the cluster.
\item The signaling server can serve as a tie-breaker. This would allow
  the remaining leader successfully commit the removal of the other
  node. In the case of a partition, the signaling server must only
  vote for one of the nodes.
\item The node with the lower ID could have an implicit tie-breaker vote.
  This would allow one node of the partition to continue in the case
  of a network partition. If a node dies or is killed, this would only
  allow the cluster to continue to operate if the node that survives
  happens to have the lower ID.
\item The system could allow manual user intervention. However, this is
  only likely to work correctly in the case where the user of the last
  remaining node can be certain that other nodes will never return
  otherwise there is a risk of data-corruption.
\end{enumerate}

Running a Raft cluster in a browser context can lead to some level of
timer instability. The JavaScript execution context is a fully event
driven environment with a single thread of execution for a given
JavaScript context. JavaScript code is run when browser (or Node.js)
events trigger JavaScript code that is registered to that event. In
order to implement logic that runs periodically, JavaScript code is
registered as a callback timer using the setTimer, setInterval or
requestAnimationFrame APIs. However, none of these mechanisms is
a precise timer. When a timer fires, the callback code they refer to
is added to the JavaScript execution queue and events on the queue are
processed in the order they arrive. This means callbacks may be
delayed by other code that is already running or queued first
\cite{resig2013secrets}.

However, there is another browser mechanism that is less well known
but that causes even larger delays in timer code execution. Modern
browsers often have dozens or even hundreds of windows and/or tabs
open simultaneously. In order to reduce CPU load, increase user
responsiveness and improve battery life, most modern browsers will
throttle the JavaScript callback timers for all windows and tabs that
are hidden or have not been directly interacted with for some period
of time. This throttling can cause timers to slow down by an order of
magnitude or more. %TODO/CITE

In the context of Raft over WebRTC, the slowdown of background page
timers means that whenever the leader node is residing on a page that
is in the background, the propagation of transaction log entries slows
down substantially. This also means that other nodes which are in the
foreground are more likely to have an election timeout timer fire and
they are more likely to become a leader because they will be able to
send and receive votes more quickly than nodes that are in the
background.  In this sense the problem is self-correcting because over
the long term because higher performing nodes more are likely to start
and win elections. This also indicates dynamic adjustment of timeout
values may be a fruitful area of further study.

In the context of the full chat application the slowdown of background
pages is even more pronounced because each addition to the chat
history can take several messages to fully propagate the new date. An
interesting area of further study would be to extend the Raft protocol
to more proactively adapt to network and node performance conditions
(delayed timers) and more intentionally transition leadership to more
capable nodes.

@@

To simplify the execution of the tests and to make the testing
infrastructure more portable, a docker container environment was used
to encapsulate and execute the SlimerJS environment. The container is
built using \emph{test/Dockerfile} which is in turn derived from
the "fentas/slimerjs" docker image \cite{fentas:slimerjs}.
This custom Dockerfile was created because the upstream version of the
Dockerfile used an older version of XULRunner that did not support
more than 7 WebRTC clients communicating together simultaneously. This
issue was discovered during testing.

@@

Three different tests were created to test the basic Raft over WebRTC
application and the chat application. Each of the tests takes
a parameter indicating the number of nodes/pages to launch and test
again. The tests scripts files are:

@@

\begin{enumerate}
\item Explore the use of alternate WebRTC transport modes for relaxing
    ordering constraints and delivery guarantees. Since the Raft
    protocol is tolerant of out-of-order and dropped messages, it is
    possible that Raft over WebRTC could be more efficient with one or
    both of these constraints relaxed.
\item Test the survivability of the cluster when the signaling server
    becomes completely unresponsive or very slow.
\item Explore models for dynamically adjusting Raft timeout values to
    account for changing network and browser conditions.
\item Explore models for dealing with the case where the cluster goes
    from two nodes down to one node. Can usefulness and availability
    be improved without sacrificing Raft safety constraints.
\item Test the system in more diverse network environments. Determine
    the reliability of establishing direct browser-to-browser
    communication.  Determine if corporate firewalls make this as
    difficult as is implied in some of the documentation and standards
    texts. %TODO/CITE
\item Perform quantitative and real-world testing to better
    characterize the performance, availability and scalability
    characteristics of the system.
\item Explore models for timing out and automatically removing nodes
    when they become unresponsive without needing to be notified by
    the signaling server that the nodes have disconnected.
\item Consider extensions to the Raft protocol that enable slow
    leaders to be proactively detected so that a more capable node
    (network, CPU, etc) can be elected. This would help address that
    background page/tag throttling of modern browsers.
\item Test larger scale deployments spanning larger and more diverse
    network domains. Determine the factors that affect scalability and
    characterize scaling limits.
\item Perform wider cross-browser testing including Internet Explorer
    and Opera.
\item Implement Raft log compaction and full client linearizability.
\item Explore models that allow for a combination of strict shared
    state and high performance bulk data such as media or object
    blobs. For example, the data with full consensus might contain
    hashes that to the refer to bulk data.
\item Implement a Node.js based Raft over WebRTC application. This
    would allow non-browser contexts to participate in the Raft
    cluster together with the browser clients. PeerJS issue \#103 must
    be resolved for this to be feasible \cite{peerjs:103}.
\item Explore a design where some nodes of the cluster are non-voting
    and are not counted as part of transaction commit quorum. However,
    these nodes would still be included in the log replication
    algorithm. One model this would enable would be a configuration
    where a small group of voting nodes serve as the hub for a larger
    group of non-voting nodes. This may enable higher scaling of the
    system. It also enables interesting applications such as a logging
    or backup nodes.
\end{enumerate}
